Leadership across cultures and contexts: an exploratory analysis

### 2020-06-18 leadership_across_cultures_contexts.Rmd
Tweaks to AIC and mst-knn wording

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since the seminal WEIRD people paper [@henrich_weirdest_2010], many social scientists have taken up the challenge to obtain data from "non-WEIRD" populations. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially more diversity within "non-WEIRD" populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
<!-- Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group contex. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2020; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2020; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2020; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020]. -->

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and leaders emerge in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2020]. Leaders also play important roles in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012].

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and nation-state organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2020]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

<!-- STATUS VS LEADERSHIP -->
Leaders have disproportionate group influence, whereas status is often conceptualized as a relative indicator of the social value of the individual by the group and of their relative access to contested resources [e.g., @blader_whats_2014; @von_rueden_roots_2014]. Leaders are expected to be high status [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014; @buss_human_2020]. The causal relationship between the two, however, is likely to be complex, context-specific, and driven by underlying correlates [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014].

<!-- MALE BIAS -->
Substantial evidence suggests a near universal male-bias in community leadership across cultures [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2020; @tiger_female_1978]. Sex, however, is confounded with factors such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2020; @von_rueden_sex_2018].

@@

In contrast to the qualities and functions dimensions, operationalizations of the costs and benefits of leadership were developed *a priori* based on theoretical literature. For both followers and leaders, we coded evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories (Table S\@ref(tab:costbenefitvars)). Some of the costs and benefits (e.g., mating) were similar to leader qualities (e.g., polygyny). The distinction is that costs and benefits were transactional whereas qualities were descriptive (e.g., leaders obtain wives vs. have wives).

@@

ZG and KS then discussed every instance of disagreement between their respective coding on all leadership dimensions and group structure variables to produce a consensus matrix, which is used in the analyses here and added to the published data repository (in a data frame named *leader_text2*). See the *Coding example* section of the supplementary information (SI).

@@

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, minimal spanning trees and nearest neighbors (mst-kNN), and hierarchical cluster analysis. See the SI for descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

Throughout, we compare regression models of the same outcome variable using the Akaike information criterion (AIC). $AIC=2k - 2 ln(\hat{L})$, where $k$ is the number of model parameters and $\hat{L}$ is the maximum likelihood. AIC penalizes models with more parameters and rewards models with better fit. Hence, a model with a lower AIC value is considered to be superior.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database of text record comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures.

@@

The majority of text records pertained to male leadership (`r male_leader_pct`%) and very few pertained to female leadership (`r female_leader_pct`%), with the remaining records not sex-specific (Table S\@ref(tab:fourfactorsumm)). There were female leaders in almost all contexts, however, except military and religious. Female leaders did tend to lead within kin and residential subgroups, though: `r female_residential_pct`% of female leaders led at the residential subgroup level, for instance, compared to `r male_residential_pct`% of male leaders. The majority of text records which exclusively described female leaders were from horticulturalists (`r female_hort_pct`%) (Table S\@ref(tab:sexsubtable)).

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (Tables S\@ref(tab:qualfuncvars) and S\@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and then computed the proportion of cultures with at least one text record providing support for that dimension (as described in the Methods section). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

@@

The random intercepts indicated that support for many variables did vary modestly by ethnographer and culture. Most of the standard deviations of the random author and culture intercepts for each model were between 0 and 1, on the scale of the linear predictor (Figures S\@ref(fig:FunctionRE) and S\@ref(fig:QualityRE)). In a few cases, one or the other was exactly 0, which can occur in models like ours that have many random effect levels (e.g., `r length(unique(all_data$author_ID))` authors) and a relatively small data set [@bates2015].

@@

```{r heatmapdrop1, fig.width=12, fig.cap="Leadership dimensions and features that, according to a decrease in AIC relative to intercept-only models, varied by group context, subsistence strategy, continental region, and/or leader sex. A. Leadership dimensions. B. Leadership features. Colors represent the relative change in AIC value after dropping that factor from a multiple logistic mixed effects regression model. Yellow: dropping factor substantially increased AIC relative to other factors. Black: dropping factor did not substantially increase AIC relative to other factors. Delta AIC values were standardized by rows, and therefore can only be compared within rows."}

@@

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, for subsistence strategy, there were three leadership dimensions in which it played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)A). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For Group context, see Figure S\@ref(fig:heatmapgroups); for Region, see Figure, S\@ref(fig:heatmapregion); and for Leader sex, see Figure S\@ref(fig:heatmapsex).

@@

The *Competencies* feature combined dimensions of prestige-style leadership, such as knowledge and experience, with dimensions of dominance-style leadership, such as aggressiveness and feared. It also included supernatural qualities. Shamans and other leaders with supernatural qualities were surprisingly common in the ethnographic record. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of *Shamanism* as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, which are a form of knowledge or expertise, and *Experienced/accomplished*). See Figure S\@ref(fig:shamanism).

@@

There were no text records that described shamans or leaders with supernatural qualities who were exclusively female. However, there were `r sum(df_shaman$demo_sex == 'both' & df_shaman$shamanism)` text records describing such leaders of both sexes. There was also more evidence of shamans and leaders with supernatual qualities in the Americas compared to other regions. See Figure \@ref(fig:shamanregion).

@@

<!-- Leadership and followership stem from biological market forces and information-processing roles [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019; @pietraszewski_evolution_2020]. Much of human leadership goes beyond group coordination problems, such as leaders important roles in devising strategic outcomes between individuals, offering advice or decision-making services to individuals, and in making direct prosocial investments that do not involve substantial coordination. We therefore suggest leadership dynamics emerge from inter-individual competencies and leaders' abilities to provide benefits *or* impose costs, with these dimensions interacting with market forces rooted in follower preferences, group context, and cultural norms.  -->

 <!-- Whereas @pietraszewski_evolution_2020 emphasizes leaders ability to propose various enterprises and followers ability to evaluate them, -->

@@

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature might correspond, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2020].

@@

When ethnographic texts described female leaders exclusively, they were relatively more likely to be described as married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits (Figure S\@ref(fig:heatmapsex)). Female leaders also had higher levels of evidence for the *Social, reproductive, and material success* "feature" variable (Figure S\@ref(fig:heatmapfeaturesex)). Evidence for female leaders was also more likely to be found in kin groups and residential subgroups. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985; @yanca_female_2004]. @garfield_investigating_2020 similarly found evidence of positive assortative mating of high status leaders in a small-scale society with relative gender-egalitarianism. Our results are also consistent with theoretical arguments emphasizing women as leaders of families [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019].

@@

In our view, social context is critical to these discussions [see @wiessner_power_2010]. Leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social "statuses." In many cultures, for example, communities form work-groups and appoint "chiefs-for-a-day" who manage the work but have little status or authority otherwise [e.g., @macfarlan_competitive_2012]. @garfield_evolutionary_2019 found that leader organization dimensions clustered separately from leader prestige dimensions, suggesting they are distinct. @garfield_investigating_2020 found that some leaders among recently settled hunter-gatherers had organizing responsibilities but unexceptional levels of respect. Some older individuals, on the other hand, were highly respected but had little social influence or authority.

@@

@henrich_evolution_2001 introduced this important distinction to social science disciplines outside of anthropology, agreeing with @tiger_dominance_1970 and Barkow [-@barkow_prestige_1975; -@barkow_prestige_1980] on a primate heritage of dominance, but offering an evolutionary account of prestige rooted in social rather than sexual selection for culturally acquired competencies. @henrich_evolution_2001 inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2020].

@@

Leaders, possibly due to their special competencies, are able to devise and implement strategic solutions to group problems including resolving conflicts within the group (the most widely supported function in our sample) and providing counsel (also widely supported), which again are functions that are commonly noted in the anthropological and psychological leadership literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002]. The feature of leader organizational functions aligns with many evolutionary theories that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; @pietraszewski_evolution_2020; for review, see @garfield_evolutionary_2019].

@@

Our revised model proposes that leaders deploy cognitive, social, material, and somatic capital to provide benefits and impose costs on followers. If their abilities to provide benefits or impose costs are similar to other groups members, as in "chiefs-for-a-day", they acquire little status, but if their abilities are exceptional relative to other group members, they gain status via biological market mechanisms [@hammerstein_biological_2016; @garfield_evolutionary_2019; @hagen_leadership_2019; @pietraszewski_evolution_2020]. There was particularly strong evidence for cognitive and social capital, with cognitive capital associated with both providing benefits and imposing costs, particularly among shaman leaders (Figure \@ref(fig:fivefold)).

@@

<!-- DIVERSE FUNCTIONS -->
<!-- Contemporary theoretical models of the evolution of leadership often emphasize leaders roles in organizing collective action and in managing between group conflicts [@hooper_theory_2010; @pietraszewski_evolution_2020; @price_evolution_2014]. Results here do support the important functions of leaders in organizing cooperation and facilitating social functions -- services which provide direct benefits to the group. However, the ubiquity of conflict resolution by leaders across social contexts and cultures calls for greater attention to the roles of leaders of all types in managing intra-group conflict between group members -- services tailored to provide direct benefits to a few individuals and more indirect group benefits [@glowacki_leadership_2015; @wiessner_collective_2019; @pupu_challenges_2018; @hagen_leadership_2019]. -->

>>>>

<!-- WEIRD PEOPLE PROBLEMS -->
There is now a broad consensus on the importance of data from the full range of human cultural diversity. Since the seminal WEIRD people paper [@henrich_weirdest_2010, *The weirdest people in the world?*], many social scientists have taken up the challenge to obtain data from "non-WEIRD" populations. Dichotomizing populations and their psychologies and social dynamics as WEIRD or non-WEIRD, however, is a mistake.^[A Google Scholar search for "non-WEIRD" produced over 1000 results.] As anthropology has conclusively demonstrated, there is enormous diversity across societies that does not remotely resemble a dichotomy.

@@

There is potentially more diversity within non-WEIRD populations as there is within WEIRD ones, and there is also substantial overlap of the behaviors of people everywhere [@brown_human_1991]. @henrich_weirdest_2010 in their critique of the over-reliance on WEIRD samples emphasized a holistic approach to cultural diversity within and across "telescoping" levels of human societies. It is therefore important to assess cultural variation without assuming that everyone within a culture is essentially the same, and without putting undue emphasis on either the between-population differences or similarities.

@@

<!-- EVOLUTIONARY THEORY APPROACH -->
<!-- Substantial scientific work across psychology, sociology, political science, and managerial studies has focused on trying to uncover interrelationships of dimensions of leadership, including the traits of leaders, their roles, and relationships with culture, economics, and group contex. Until recently, this large body of work was relatively disconnected from approaches in anthropology and biology. Under a unifying evolutionary framework, social and biological scientists have now produced a sizable body of theory on the evolution and cross-cultural patterning of leadership and followership [e.g., @price_evolution_2014; @henrich_big_2015; @richerson_tribal_2009; @van_vugt_evolutionary_2006; @pietraszewski_evolution_2019; @hooper_theory_2010; @kaplan_evolutionary_2009; @von_rueden_roots_2014; @hagen_leadership_2019], accompanied by a growing body of empirical results [e.g., @bowser_womens_2010; @alami_mothers_2020; @cheng_two_2013; @wiessner_collective_2019; @redhead_dynamics_2018; @garfield_investigating_2019; @garfield_evolutionary_2019; @von_rueden_leadership_2014; @von_rueden_sex_2018; @smith_leadership_2016; @smith_obstacles_2018; @lonati_what_2019]. This integration as been informative [see @van_vugt_genes_2020]. -->

@@

## The functions of leadership across cultures and contexts
<!-- INTRO -->
@roscoe_social_2009 proposed leadership in small-scale societies is structured by the joint demands of facilitating within-group conflict resolution and cooperation, and signaling coalitional strength to rival groups. Among many small-scale societies, which generally lack institutionalized political structures, leadership at the community level is typically informal and leaders emerge in response to group demands [@von_rueden_leadership_2015; @kantner_identifying_2010; @lewis_leaders_1974; @boehm_egalitarian_1993]. Theoretical literature suggests these demands most commonly arise in the contexts of collective actions, resolving conflicts between families, and punishing norm violations [e.g., @glowacki_leadership_2015; @gavrilets_convergence_2016; @hooper_theory_2010; @von_rueden_solving_2015; @henrich_big_2015; @pietraszewski_evolution_2019]. Leaders also play important roles in ritual contexts, such as in initiations [@singh_cultural_2017], facilitating marriages [@walker_evolutionary_2011], and organizing feasts [@wiessner_food_1996]. Between-group interactions, including both conflict and cooperation, are also common among hunter-gatherers and other nonindustrial societies [@glowacki_evolutionary_2017; @hames_pacifying_2019; @richerson_cultural_2016; @henrich_evolution_2009; @apicella_social_2012].

@@

<!-- MORE UNIVERSAL QUALITIES FROM WEIRD AND NON WEIRD -->
Intelligent and knowledgeable individuals [e.g., @johnson_leadership_2015; @judge_intelligence_2004; @neel_being_1980; @antonakis_does_2009], elders, and respected individuals tend to have greater influence over community decisions across cultures, including among egalitarian hunter-gatherers, stratified chiefdoms, and nation state organizations [@silverman_how_1978; @tooby_cognitive_2006; @garfield_review_2019; @bass_handbook_1990; @garfield_evolutionary_2019; @garfield_investigating_2019]. Leaders also tend to have large social networks [@von_rueden_leadership_2014; @von_rueden_roots_2014; @walker_living_2012], and embody the qualities most respected by the group [@henrich_evolution_2001; @van_vugt_evolutionary_2006; @collier_politics_1981; @low_men_1992; @roscoe_social_2019]. Leaders across multiple contexts tend to be taller and more physically formidable, which might facilitate resolving inter- and intra-group  conflicts [@lukaszewski_role_2016; @von_rueden_leadership_2014; @von_rueden_roots_2014]. Charisma and oratory skills are common properties of leaders when they must rely on persuasion to influence community members [@grabo_charismatic_2016].

<!-- STATUS VS LEADERSHIP -->
Leaders have disproportionate group influence, whereas status is often conceptualized as a relative indicator of the social value of the individual by the group and of their relative access to contested resources [e.g., @blader_whats_2014; @von_rueden_roots_2014]. Leaders are expected to be high status [e.g., @von_rueden_leadership_2015; @van_vugt_genes_2020; @cheng_dominance_2019; @cheng_psychology_2014]. The causal relationship between the two, however, is likely to be complex, context-specific, and driven by underlying correlates [e.g., @blader_whats_2014; @cheng_dominance_2019; @kantner_identifying_2010; @von_rueden_roots_2014].

<!-- MALE BIAS -->
Substantial evidence suggests a near universal male-bias in community leadership across human societies [@low_sex_1992; @whyte_status_1978; @sanday_female_1981], as well as across diverse nonhuman social species [@smith_obstacles_2018; @tiger_female_1978]. Sex, however, is confounded with factors such as social network size, economic specialization, and education, that are also associated with leadership. After controlling for these factors, some studies have found that sex is not a strong predictor of community leadership [@yanca_female_2004; @garfield_investigating_2019; @von_rueden_sex_2018].

@@

In contrast to the qualities and functions dimensions, operationalizations of the costs and benefits of leadership were developed *a priori* based on theoretical literature. For both followers and leaders, we coded evidence of costs paid or benefits received in the following categories: social status, reduced or increased risk of harm or conflict, social services, territory, mating, fitness, food resources, material resources, and other miscellaneous costs or benefits not captured by these categories. See Table S\@ref(tab:costbenefitvars). Some of the costs and benefits (e.g., mating) were similar to leader qualities (e.g., polygyny). The distinction is that costs and benefits were transactional whereas qualities were descriptive (e.g., leaders obtain wives vs. have wives).

@@

ZG and KS then discussed every instance of disagreement between their respective coding on all leadership dimensions and group structure variables to produce a consensus matrix, which is used in the analyses here and added to the published data repository (in a data frame named *leader_text2*). For an illustration of the coding process, see the *Coding example* section of the SI.

@@

Our fourth goal was to explore lower dimensional representations of these data that might reveal informative "features." These included logistic PCA, minimal spanning trees and k-nearest neighbors (mst-kNN), and hierarchical cluster analysis. See the supplementary information (SI) for brief descriptions of these methods. Our fifth goal was to assess variation in our lower-dimensional "features" by group context, subsistence strategy, continental region, and leader sex, using logistic mixed effects models fit with the lme4 package. Our sixth goal was to evaluate the features in light of influential evolutionary theories of leadership. Drawing on the empirically derived features as a guide, we aimed to construct "elements" of leadership that corresponded to elements of influential theories, examining intersections of these elements across all text records.

@@

Throughout, we compare regression models of the same outcome variable using the Akaike information criterion (AIC). $AIC=2k - 2 ln(\hat{L})$, where $k$ is the number of model parameters and $\hat{L}$ is the maximum likelihood. AIC increases as model complexity increases, and decreases with improved fit. Hence, a model with a lower AIC value is considered to be superior.

@@

The geographic distribution of the cultures in our sample is displayed in Figure \@ref(fig:geographic) (see also Figure S\@ref(fig:subsisttree)). The database of text record comprised `r nrow(all_data)` text records from `r length(unique(leadershipdata::documents$d_ID))` unique documents describing `r length(unique(all_data$d_culture))` cultures. The number of text records per culture ranged from `r min(leader_cult$number_leader_records, na.rm=T)` to `r max(leader_cult$number_leader_records, na.rm=T)`, with a median of `r median(leader_cult$number_leader_records, na.rm=T)`, a mean of `r signif(mean(leader_cult$number_leader_records, na.rm=T), 3)`, and a standard deviation of `r signif(sd(leader_cult$number_leader_records, na.rm=T), 3)`. The text records were generally short, ranging from `r textstats$min` to `r textstats$max` words, with a median word count of `r textstats$median`, a mean of `r textstats$mean`, and standard deviation of `r textstats$sd`. Of the `r nrow(all_data)` text records, `r final_record_count` had evidence for at least one dimension in the study.

@@

The majority of text records pertained to male leadership (`r male_leader_pct`%) and very few pertained to female leadership (`r female_leader_pct`%), with the remaining records not sex-specific. See Table S\@ref(tab:fourfactorsumm). There were female leaders in almost all contexts, however, except military and religious. Female leaders did tend to lead within kin and residential subgroups, though: `r female_residential_pct`% of female leaders led at the residential subgroup level, for instance, compared to `r male_residential_pct`% of male leaders. The majority of text records which exclusively described female leaders were from horticulturalists (`r female_hort_pct`%). See Table S\@ref(tab:sexsubtable).

@@

For each of the `r length(c(quality_vars, function_vars, leader_cost_vars, leader_benefit_vars, follower_benefit_vars, follower_cost_vars))` coded dimensions (see Tables S\@ref(tab:qualfuncvars) and S\@ref(tab:costbenefitvars)), we computed the proportion of text records (95% CI) that provided support and then computed the proportion of cultures with at least one text record providing support for that dimension (as described in the Methods section). See Figure \@ref(fig:varsupport). For a table of all coefficients and 95% CIs, see Table S\@ref(tab:glmmcoefs).

@@

The random intercepts indicated that support for many variables did vary modestly by ethnographer and culture. Most of the standard deviations of the random author and culture intercepts for each model were between 0 and 1 (on the scale of the linear predictor). See Figures S\@ref(fig:FunctionRE) and S\@ref(fig:QualityRE). In a few cases, one or the other was exactly 0, which can occur in models like ours that have many random effect levels (e.g., `r length(unique(all_data$author_ID))` authors) and a relatively small data set [@bates2015].

@@

```{r heatmapdrop1, fig.width=12, fig.cap="Leadership dimensions and features that, according to a decrease in AIC relative to intercept-only models, varied by group context, subsistence strategy, continental region, and/or leader sex. A. Leadership dimensions. B. Leadership features. Colors represent the relative change in AIC value after dropping that factor from a multiple logistic mixed effects regression model. Yellow: dropping factor substantially increased AIC relative to other factors. Black: dropping factor did not substantially increase AIC relative to other factors. For display purposes, delta AIC values were standardized by rows, and therefore can only be compared within rows."}

@@

For each of our four factors, we then investigated which levels of the factor was associated with higher or lower levels of evidence for the leadership dimension. For instance, there were three leadership dimensions in which subsistence strategy played an important role: *No coercive authority*, *Provide subsistence*, and *Distribute resources* (See Figure \@ref(fig:heatmapdrop1)A). Hunter-gatherers had particularly high levels of evidence for *No coercive authority*, relative to the other subsistence strategies (and averaging over the levels of the other three factors), and Pastoralists had relatively low evidence for *Distribute resources* (see Figure S\@ref(fig:heatmapsubsis)). For Group context, see Figure S\@ref(fig:heatmapgroups); for Region, see Figure, S\@ref(fig:heatmapregion); and for Leader sex, see Figure S\@ref(fig:heatmapsex).

@@

The *Competencies* feature combined dimensions of prestige-style leadership, such as knowledge and experience, with dimensions of dominance-style leadership, such as aggressiveness and feared. It also included supernatural qualities. The MST-kNN analysis similarly clustered leaders with ritual and medicinal functions, which rely on special knowledge, with those who had supernatural, feared or killer qualities (see Figure S\@ref(fig:mstknn)).

Shamans and other leaders with supernatural qualities were surprisingly common in the ethnographic record. Of our `r nrow(leader_text2)` text records, `r sum(df_shaman$shaman)` used the word "shaman", `r supernatural_n` provided evidence for supernatural leader qualities, and `r sum(df_shaman$shamanism)` provided evidence for one or both (`r shamanism_pct`% of all text records). We term the latter variable *Shamanism*. We fit a logistic elasticnet model of *Shamanism* as a function of all leadership dimensions (minus *Supernatural*). This model revealed that shaman leaders combined dimensions of dominant leaders (*Feared*) with dimensions of prestigious leaders (*Medical functions*, which are a form of knowledge or expertise, and *Experienced/accomplished*). See Figure S\@ref(fig:shamanism).

@@

There were no text records that described shamans or leaders with supernatural qualities who were exclusively female. However, there were `r sum(df_shaman$demo_sex == 'both' & df_shaman$shamanism)` text records describing such leaders of both sexes. There was also more evidence of shamans and leaders with supernatual qualities in the Americas compared to other regions. See Figure S\@ref(fig:shamanregion).

@@

<!-- Leadership and followership stem from biological market forces and information-processing roles [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019; @pietraszewski_evolution_2019]. Much of human leadership goes beyond group coordination problems, such as leaders important roles in devising strategic outcomes between individuals, offering advice or decision-making services to individuals, and in making direct prosocial investments that do not involve substantial coordination. We therefore suggest leadership dynamics emerge from inter-individual competencies and leaders' abilities to provide benefits *or* impose costs, with these dimensions interacting with market forces rooted in follower preferences, group context, and cultural norms.  -->

 <!-- Whereas @pietraszewski_evolution_2019 emphasizes leaders ability to propose various enterprises and followers ability to evaluate them, -->

@@

These potentially universal dimensions of leadership correspond to those highlighted by several evolutionary theories that have drawn heavily on select ethnographic cases [@garfield_review_2019]. They partially correspond to the prestige model [@henrich_big_2015; @henrich_evolution_2001], for example, which emphasizes knowledge, skill and generosity, but they do not include a key feature of that model: emulation of prestigious leaders. We labeled one of our leader quality features "cultural conformity", however, in which leaders are culturally conservative and exemplify desirable traits such as honesty, humility, and "proper behavior" (Figure \@ref(fig:cluster)). This feature might correspond, at least in part, to leaders who serve as models for their followers [see also @garfield_investigating_2019].

@@

When ethnographic texts described female leaders exclusively, they were relatively more likely to be described as married to a high-quality spouse in a polygynous marriage, and the recipients of various mating benefits (Figure S\@ref(fig:heatmapsex)). Female leaders also had higher levels of evidence for the *Social, reproductive, and material success* "feature" variable (Figure S\@ref(fig:heatmapfeaturesex)). Evidence for female leaders was also more likely to be found in kin groups and residential subgroups. This suite of characteristics is consistent with discussions of high status first-wives who gain and maintain increased social influence across the lifespan by leveraging their extended kin and social networks [@goodale_tiwi_1971; @brown_her_1985; @yanca_female_2004]. @garfield_investigating_2019 similarly found evidence of positive assortative mating of high status leaders in a small-scale society with relative gender-egalitarianism. Our results are also consistent with theoretical arguments emphasizing women as leaders of families [e.g., @garfield_evolutionary_2019; @hagen_leadership_2019].

@@

In our view, social context is critical to these discussions [see @wiessner_power_2010]. Leadership and followership emerge across levels of social organization and individuals simultaneously maintain multiple social "statuses." In many cultures, for example, communities form work-groups and appoint "chiefs-for-a-day" who manage the work but have little status or authority otherwise [e.g., @macfarlan_competitive_2012]. @garfield_evolutionary_2019 found that leader organization dimensions clustered separately from leader prestige dimensions, suggesting they are distinct. @garfield_investigating_2019 found that some leaders among recently settled hunter-gatherers had organizing responsibilities but unexceptional levels of respect. Some older individuals, on the other hand, were highly respected but had little social influence or authority.

@@

@henrich_evolution_2001 introduced this important distinction to social science disciplines outside of anthropology, agreeing with @tiger_dominance_1970 and Barkow [-@barkow_prestige_1975; -@barkow_prestige_1980] on a primate heritage of dominance, but offering an evolutionary account of prestige rooted in social rather than sexual selection for culturally acquired competencies. @henrich_evolution_2001 inspired much subsequent work across the behavioral sciences [e.g., @van_vugt_dual_2019; @cheng_two_2013; @maner_2016; @halevy_status_2012; @suessenbach_dominance_2019; @maner_dominance_2017; @garfield_evolutionary_2019; @garfield_investigating_2019].

@@

Leaders, possibly due to their special competencies, are able to devise and implement strategic solutions to group problems including resolving conflicts within the group (the most widely supported function in our sample) and providing counsel (also widely supported), which again are functions that are commonly noted in the anthropological and psychological leadership literature [@wiessner_risk_1982; @wiessner_collective_2019; @boehm_hierarchy_1999; @knauft_violence_1991; @de_cremer_intergroup_2002]. The feature of leader organizational functions aligns with many evolutionary theories that emphasize the role of leaders in organizing collective actions [e.g., @hooper_theory_2010; @van_vugt_cognitive_2007; @glowacki_formation_2016; @pietraszewski_evolution_2019; for review, see @garfield_evolutionary_2019].

@@

Our revised model proposes that leaders deploy cognitive, social, material, and somatic capital to provide benefits and impose costs on followers. If their abilities to provide benefits or impose costs are similar to other groups members, as in "chiefs-for-a-day", they acquire little status, but if their abilities are exceptional relative to other group members, they gain status via biological market mechanisms [@hammerstein_biological_2016; @garfield_evolutionary_2019; @hagen_leadership_2019; @pietraszewski_evolution_2019]. There was particularly strong evidence for cognitive and social capital, with cognitive capital associated with both providing benefits and imposing costs, particularly among shaman leaders (Figure \@ref(fig:fivefold)).

@@

<!-- DIVERSE FUNCTIONS -->
<!-- Contemporary theoretical models of the evolution of leadership often emphasize leaders roles in organizing collective action and in managing between group conflicts [@hooper_theory_2010; @pietraszewski_evolution_2019; @price_evolution_2014]. Results here do support the important functions of leaders in organizing cooperation and facilitating social functions -- services which provide direct benefits to the group. However, the ubiquity of conflict resolution by leaders across social contexts and cultures calls for greater attention to the roles of leaders of all types in managing intra-group conflict between group members -- services tailored to provide direct benefits to a few individuals and more indirect group benefits [@glowacki_leadership_2015; @wiessner_collective_2019; @pupu_challenges_2018; @hagen_leadership_2019]. -->
